Identifying supermassive black hole recoil in elliptical galaxies

### 2024-10-10 papers/paper-corekick/paper-corekick.tex
Bring changes from Overleaf to git

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
%\usepackage{amssymb}	% Extra maths symbols
\usepackage{bm}         % Bold math fonts

@@

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%
% drafting macros
\usepackage{xcolor}
\newcommand{\drafting}[1]{
{\leavevmode\color[RGB]{224, 77, 24}#1}
}

@@

\newcommand{\Msun}{\ensuremath{\mathrm{M}_{\sun}}}            % Solar Mass
\newcommand{\kmps}{\ensuremath{\mathrm{km}\,\mathrm{s}^{-1}} }% km/s formatting
\newcommand{\Reff}{\ensuremath{R_\mathrm{e}}}                 % effective radius
\newcommand{\rb}{\ensuremath{r_\mathrm{b}}}                   % core radius
\newcommand{\vk}{\ensuremath{v_\mathrm{kick}}}                % kick velocity

\newcommand{\dd}[1]{\ensuremath{\mathrm{d}#1}}                % integral dx
\newcommand{\dv}[2]{\ensuremath{\frac{\dd{#1}}{\dd{#2}}}}     % derivative
\newcommand{\dnv}[3]{\ensuremath{\frac{\mathrm{d}^#1#2}{\dd{#3}^#1}}}  % nth derivative
\newcommand{\vb}[1]{\ensuremath{\bm{#1}}}                     % vectors

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Cores from kicks]{Cores from kicks: inferring the gravitational recoil history of SMBHs in elliptical galaxies.}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[A. Rawlings et al.]{
Alexander Rawlings,$^{1}$\thanks{E-mail: alexander.rawlings@helsinki.fi}
et al.
\vspace*{0.1cm}\\%
% List of institutions
$^{1}$
Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland
\\%
%$^{2}$
%Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzchild-Str 1, D-85748 Garching, Germany
}

@@

% Enter the current year, for the copyright statements etc.
\pubyear{2023}

@@

% Abstract of the paper
\begin{abstract}
\drafting{We answer the questions:
\begin{enumerate}
    \item How does a GW kick influence the stellar mass distribution?
    \item Can we infer if a GW kick has occurred in a ETG?
\end{enumerate}
}
\end{abstract}

@@

% INTRODUCTION
\section{Introduction}
There is much evidence for the presence of supermassive black holes (SMBHs) at the centres of massive galaxies, with the shadow of one such SMBH being directly imaged \citep{eht2019}.
The dynamical interactions of two or more galaxies, during a galaxy merger, are thus also expected to induce an interaction of the residing SMBHs.
The interaction of two SMBHs during the merger of two galaxies is well understood to occur in a three stage process \citep{begelman1980}.
First, as the galaxies begin to dynamically interact and the SMBHs traverse through a dense stellar field, the overdensity of stars in the wakes of the SMBHs provides a restoring force which brings the SMBHs to the centre of the two interacting galaxies \citep{chandrasekhar1943}, generally torquing the SMBHs to highly radial orbits in the process. 
Second, after the SMBHs have formed a bound binary, iterative interactions with the surrounding stars ejects stars with high velocity \citep[e.g.][]{hills1980,hills1983,quinlan1996}, thus removing orbital energy and angular momentum from the SMBH binary orbit.
The removal of stars from the central regions by this slingshot mechanism produces a `core' \citep[e.g.][]{lauer1983,lauer1985,kormendy1984}, or depletion, in the luminosity profile of the remnant galaxy \citep[e.g.][]{begelman1980,hills1983,quinlan1996,rantala2018}, and can extend up to some kiloparsecs \citep{postman2012}.
Finally, at very small, sub-parsec, separations the SMBH binary loses its remaining orbital energy and angular momentum through gravitational wave (GW) emission, driving the SMBH binary to coalescence \citep{peters1963,peters1964}.

The final GW emission from the coalescing SMBH binary occurs anisotropically, and carries with it linear momentum from the system \citep[e.g.][]{gonzalez2007}.
This results in the remnant SMBH recoiling with some velocity directed opposite to the linear momentum of the GW emission \citep{bekenstein1973}, and is termed the kick velocity. 
The kick velocity is dependent on the mass ratio between the SMBHs prior to coalescence, and the magnitude and direction of angular momentum (spin) vector of each SMBH. 
As reported by \citet{campanelli2007}, symmetries in the masses or spins of the SMBHs suppresses the resulting recoil kick imparted to the coalesced SMBH.
In particular, it is found that asymmetry in the spins of the SMBHs has a larger impact on the magnitude of the recoil kick than asymmetry in the masses \citep{campanelli2007}, with antialigned spins in the orbital plane producing the largest recoil kick velocities \citep{gonzalez2007b,tichy2007}.
Numerical relativity studies indicate that while the majority of recoil kicks are of the order of a few hundred kilometres per second, in special configurations the SMBH may be imparted a kick velocity in excess of a $2000\,\kmps$, even up to $4000\,\kmps$ \citep{campanelli2007,gonzalez2007b,tichy2007}.
Taking the central escape velocity of a typical early type galaxy (ETG) as $\sim2000\,\kmps$, there may be a non-negligible fraction of ETGs with a SMBH that has escaped the galaxy \citep{madau2004}, though not so many so as to introduce considerable scatter into the observed relation between SMBH mass and stellar velocity dispersion \citep{volonteri2007}.
In cases where the kick velocity is less than the escape velocity of the inner regions of the host galaxy, the radial oscillations of the remnant SMBH as it returns to the centre of the merger remnant is expected to influence the surrounding stellar environment. \drafting{What sort of influences? References?}

Numerous observational studies have also hinted at the existence of recoiling SMBHs, generally through either an offset from the centre of the host galaxy in velocity, in position, or both.
One of the earliest observations that pointed to a recoiling SMBH was of the quasar SDSSJ0927+2943, reported by \citet{komossa2008}, where an observed velocity offset suggested a recoil kick in excess of $2600\,\kmps$.
Another recoiling SMBH candidate was identified by an offset in velocity by \citet{steinhardt2012}, with an estimated recoil velocity greater than $4000\,\kmps$.
Other observations by \citet{comerford2014} and \citet{pesce2018} have also reported potential recoiling SMBHs, albeit at much lower kick velocities (some few hundreds of $\kmps$).
\drafting{In \citet{pesce2018}, they talk about a peculiar velocity of 70ish km/s, whereas in later work, they stated velocity is 4000 km/s. Not sure which is correct, that difference seems suspicious\dots}
Recoiling SMBHs may also be identified through a spatial offset from the host centre, where the offset can range from some parsecs \citep{batcheldor2010,lena2014,barrows2016} to over a kiloparsec \citep{koss2014,skipper2018}.
Finally, some candidate recoiling SMBHs have both spatial and velocity offsets, such as CXO J101527.2+625911 \citet{kim2017} and 2MASX J00423991+3017515 \citet{hogg2021}.

In this work, we first aim to determine how the velocity with which the SMBH is kicked affects the surrounding stellar environment.
We then aim to determine if there is an observational signature of the SMBH recoil kick that may be used to infer the occurrence of a kicked SMBH. 
\drafting{Previous work by \dots indicates \dots}

\drafting{This paper is divided as follows:}

@@

\subsection{Simulation Code}
\drafting{Taken from eccentricity paper, edited a little bit but need to ensure it is different enough.}
To investigate the effect of the recoiling SMBH on the host galaxy, and potential observational signatures of this, we run a number of numerical simulations of a galaxy merger setting using our new version of the \ketju{} code \citep{mannerkoski2023,rantala2017} coupled with \gadget{}.
\ketju{} integrates the dynamics of SMBHs, and stars in a small region thrice the stellar softening length around them, to high accuracy using the algorithmically regularised integrator \textsc{mstar} \citep{rantala2020}.
The dynamics of stellar particles beyond this small region of high integration accuracy, and all dark matter (DM) particles, is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. 
Additionally, we use hierarchical time integration, which allows for symmetric interactions and manifest momentum conservation. 
\ketju{} also includes post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014} and the fitting formula of \citet{zlochower2015} for recoil kick velocity, making \ketju{} a particularly well-suited code for investigating the consequence of SMBH recoil self-consistently in a galaxy merger environment.

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
%\usepackage{amssymb}	% Extra maths symbols
\usepackage{bm}         % Bold math fonts
\usepackage{pifont}

@@

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%
% drafting macros
\usepackage{xcolor}
\newcommand{\alex}[1]{
{\leavevmode\color[RGB]{224, 77, 24}Alex: #1}
}
\newcommand{\atte}[1]{{\color{teal}Atte: #1}}
\newcommand{\peter}[1]{{\color{purple}Peter: #1}}
\newcommand{\maxx}[1]{{\color{orange}Max: #1}}
\newcommand{\ruby}[1]{{\color{blue}Ruby: #1}}
\newcommand{\sonja}[1]{{\color{violet}Sonja: #1}}

@@

\newcommand{\Msun}{\ensuremath{\mathrm{M}_{\sun}}}            % Solar Mass
\newcommand{\Lsun}{\ensuremath{\mathrm{L}_{\sun}}}            % Solar luminosity
\newcommand{\kmps}{\ensuremath{\mathrm{km}\,\mathrm{s}^{-1}} }% km/s formatting
\newcommand{\Reff}{\ensuremath{R_\mathrm{e}}}                 % effective radius
\newcommand{\rb}{\ensuremath{r_\mathrm{b}}}                   % core radius
\newcommand{\rbbin}{\ensuremath{r_\mathrm{b,0}}}                   % initial core radius
\newcommand{\vk}{\ensuremath{v_\mathrm{kick}}}                % kick velocity
\newcommand{\vkmax}{\ensuremath{v_\mathrm{kick, max}}}        % max kick velocity

\newcommand{\dd}[1]{\ensuremath{\mathrm{d}#1}}                % integral dx
\newcommand{\dv}[2]{\ensuremath{\frac{\dd{#1}}{\dd{#2}}}}     % derivative
\newcommand{\dnv}[3]{\ensuremath{\frac{\mathrm{d}^#1#2}{\dd{#3}^#1}}}  % nth derivative
\newcommand{\vb}[1]{\ensuremath{\bm{#1}}}                     % vectors
\newcommand{\ordinal}[1]{\ensuremath{#1^\mathrm{th}}}                     % ordinal th
\newcommand{\ordinalthird}[1]{\ensuremath{#1^\mathrm{rd}}}                     % ordinal rd

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Identifying SMBH recoil]{
Identifying supermassive black hole recoil in elliptical galaxies
}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[A. Rawlings et al.]{Alexander Rawlings,$^{1}$\thanks{E-mail: alexander.rawlings@helsinki.fi}
Atte Keitaanranta,$^{1}$
Max Mattero,$^{1}$
Sonja Soininen,$^{1}$
Ruby J. Wright,$^{2,1}$
\newauthor
Noa Kallioinen,$^{3}$
Shihong Liao,$^{4}$
Antti Rantala,$^{5}$
Peter H. Johansson,$^{1}$
Thorsten Naab,$^{5}$ and
\newauthor Dimitrios Irodotou$^{6}$
\vspace*{0.1cm}\\
% List of institutions
$^{1}$
Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland
\\
$^{2}$
International Centre for Radio Astronomy Research (ICRAR), University of Western Australia, Crawley, WA 6009, Australia
\\
$^{3}$
Department of Computer Science, Aalto University, Espoo, Finland
\\
$^{4}$
Key Laboratory for Computational Astrophysics, National Astronomical Observatories, Chinese Academy of Sciences, Beijing 100101, China
\\
$^{5}$
Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzchild-Str 1, D-85748 Garching, Germany
\\
$^{6}$
The Institute of Cancer Research, London SW7 3RP, UK
\\
}

@@

% Enter the current year, for the copyright statements etc.
\pubyear{2024}

@@

% Abstract of the paper
\begin{abstract}
\alex{We have 250 words to use here. Currently 247 excluding the 3 math inlines.}

We study the contribution to stellar core growth in simulations of merging massive ($M_\star>10^9\,\Msun$) elliptical galaxies by a coalesced supermassive black hole (SMBH) displaced by gravitational wave (GW) induced recoil velocity.
With controlled, dense sampling of the SMBH recoil velocity, we find the stellar core originally formed by SMBH binary scouring can grow by a factor of 2-3 with increasing recoil velocity, and a similar growth in the central mass deficit.
Using Bayesian inference we predict the distribution of stellar core sizes formed through this process to peak at $\sim1\,\mathrm{kpc}$.
An orbital decomposition of stellar particles within the core reveals that when the recoil velocity exceeds the velocity dispersion of the core, radial orbits dominate over tube orbits, providing a mass 10\% of the coalesced SMBH mass, whereas tube orbits dominate for the lowest recoil kicks.
A change in orbital structure is reflected in the anisotropy parameter, with a tangential bias present only for kick velocities less than the local stellar velocity dispersion.
With mock integral field unit observations of the stellar line of sight velocity distribution emulating the capabilities of current detectors, we observe a distinct signature in the Gauss-Hermite symmetric deviation coefficient $h_4$ that allows for an estimate of the core size due to binary scouring, irrespective of later evolution due to SMBH recoil.
Our results provide a method to estimate the SMBH recoil magnitude from observations of local elliptical galaxies, and implies these galaxies primarily experienced recoil velocities less than the stellar velocity dispersion of the core.
\end{abstract}

@@

% INTRODUCTION
\section{Introduction}
%There is much evidence for the presence of supermassive black holes (SMBHs) at the centres of massive galaxies, with the shadow of one such SMBH being directly imaged \citep{eht2019}.
%As two galaxies dynamically interact during a galaxy merger, so too is it expected that the residing SMBHs will dynamically interact.
Observations have revealed a clear bimodality in the properties of massive elliptical galaxies, with brighter ellipticals $(M_{B}\lesssim -20.5)$ typically having `cored' flat central surface brightness profiles, boxy isophotes and showing little or no rotation (slow rotators). Fainter intermediate-mass ellipticals $(-20.5 \lesssim M_{B} \lesssim -18.5)$ on the other hand typically show steeper `cuspy' central brightness profiles, more disc-like isophotes and evidence for more rotational support (fast rotators) (e.g \citealt{Kormendy1996,Faber1997,emsellem2007,Kormendy2009,Thomas2014,dullo2014,dullo2019,Quenneville2024}).

The observed bimodality can be linked to the distinct formation paths of elliptical galaxies. Intermediate-mass ellipticals have likely formed predominantly through in-situ star formation and in mergers of gas-rich disc-dominated galaxies accompanied by merger-induced starbursts resulting in cuspy central stellar profiles (e.g. \citealt{Barnes1996,Cappellari2007,hopkins2009a,hopkins2009b,Johansson2009,Krajnovic2011,Cappellari2016,Lahen2018,Forster2020}). 
More massive elliptical galaxies are on the other hand believed to have assembled in a two-stage process, in which the early assembly at redshifts of $z \gtrsim 1.5$ is dominated by in-situ star formation fuelled by cold gas flows in massive dark matter (DM) haloes, and the accretion of multiple star-bursting progenitors. The later evolution of the galaxies then proceeds mainly through gas-poor (dry) merging in which the galaxy predominantly accretes stars formed  outside the main galaxy (e.g. \citealt{Naab2009,Bezanson2009,Oser2010,Johansson2012,Wellons2015,Rodriguez-Gomez2016,Qu2017,lagos2022,Cannarozzo2023}, see also \citealt{naab2017} for a review). However, if massive galaxies have formed by merging of fainter cuspy galaxies they should also possess central cusps, as simulations have shown that mergers tend to preserve the density cusps of merging galaxies \citep{boylan2004,Dehnen2005}.

In the case of two massive elliptical galaxies undergoing a dry merger, the leading mechanism for stellar core formation is through the interaction of binary supermassive black holes (SMBHs) in the galaxy merger remnant. The merging of SMBHs 
is generally understood in the context of a three stage process \citep{begelman1980}. First, the dynamical friction from the mass overdensity in the wakes of the SMBHs provides a restoring force which brings the SMBHs to the centre of the two interacting galaxies \citep{chandrasekhar1943}, generally torquing the SMBHs to highly radial orbits in the process. Second, after the mass enclosed within the orbit of the two SMBHs is comparable to the sum of the SMBH masses, the SMBHs form a bound binary \citep{merritt2005}. Iterative interactions of the SMBHs with surrounding stars on low angular momentum orbits ejects these stars with high velocity \citep[e.g.][]{hills1980,hills1983,quinlan1996}, thus removing orbital energy and angular momentum from the SMBH binary orbit.
The removal of stars from the central regions by this slingshot mechanism produces a `core' \citep[e.g.][]{lauer1983,lauer1985,kormendy1984}, or depletion, in the luminosity profile of the remnant galaxy \citep[e.g.][]{begelman1980,hills1983,quinlan1996,rantala2018,rantala2024,nasim2021b,khonji2024,partmann2024}, that can in extreme cases extend up to $3\,\mathrm{kpc}$ \citep{postman2012}.
Finally, at very small, sub-parsec separations the SMBH binary loses its remaining orbital energy and angular momentum through gravitational wave (GW) emission, driving the SMBH binary to coalescence \citep{peters1963,peters1964}. The GW emission from such SMBH coalescence events are prime observational targets for ongoing pulsar timing array (PTA) detectors for SMBHs with masses of $M_\bullet \gtrsim 10^9\,\Msun$ \citep{agazie2023, 2023Antoniadis, 2023Xu, zic2023}, and also the forthcoming Laser Interferometer Space Antenna (LISA) mission for SMBHs with slightly lower masses of $M_\bullet \lesssim 10^8\,\Msun$ \citep{amaro-seoane2023}.

The final GW emission from the coalescing SMBH binary occurs anisotropically, and carries with it linear momentum from the system \citep[e.g.][]{gonzalez2007}.
This results in the remnant SMBH recoiling with some velocity directed opposite to the linear momentum of the GW emission \citep{bekenstein1973}, and is termed the kick, or recoil, velocity\footnote{We use the terms `kick' and `recoil' interchangeably in this work.}. 
The kick velocity is dependent on the mass ratio between the SMBHs prior to coalescence, and the magnitude and direction of the angular momentum (spin) vector of each SMBH. 
As reported by \citet{campanelli2007}, symmetries in the masses or spins of the SMBHs suppresses the resulting recoil kick imparted to the coalesced SMBH.
In particular, it is found that asymmetry in the spins of the SMBHs has a larger impact on the magnitude of the recoil kick than asymmetry in the masses \citep{campanelli2007}, with spins which are anti-aligned in the orbital plane and maximal producing the largest recoil kick velocities \citep{gonzalez2007b,tichy2007}.
Numerical relativity studies indicate that while the majority of recoil kicks are of the order of a few hundred kilometres per second, in special configurations the SMBH may be imparted a kick velocity in excess of $2000\,\kmps$, even up to $4000\,\kmps$ \citep{campanelli2007,gonzalez2007b,tichy2007}.
Taking the central escape velocity of a typical massive elliptical galaxy as $\sim2000\,\kmps$, there may be a non-negligible fraction of elliptical galaxies with an SMBH that has escaped the galaxy (e.g. \citealt{madau2004,mannerkoski2022}), though not so many so as to introduce considerable scatter into the observed relation between SMBH mass and stellar velocity dispersion \citep{volonteri2007}.
This is in clear tension with all current observations indicating that no massive elliptical galaxy in the local Universe is without an SMBH, and demands further investigation.

Numerous observational studies have also hinted at the existence of recoiling SMBHs, generally through either an offset from the centre of the host galaxy in velocity, in position, or both. One of the earliest observations that pointed to a recoiling SMBH was of the quasar SDSSJ0927+2943, reported by \citet{komossa2008}, where an observed velocity offset suggested a recoil kick in excess of $2600\,\kmps$.
Another recoiling SMBH candidate was identified by an offset in velocity by \citet{steinhardt2012}, with an estimated recoil velocity greater than $4000\,\kmps$.
Other observations by \citet{comerford2014} and \citet{pesce2018, pesce2021} have also reported potential recoiling SMBHs, albeit at much lower kick velocities (some tens to few hundreds of $\kmps$).
Recoiling SMBHs may also be identified through a spatial offset from the host centre, where the offset can range from some parsecs \citep{batcheldor2010,lena2014,barrows2016} to over a kiloparsec \citep{koss2014,skipper2018}.
Additionally, some candidate recoiling SMBHs have both spatial and velocity offsets, such as CXO J101527.2+625911 \citep{kim2017} and 2MASX J00423991+3017515 \citep{hogg2021}.
Attributing an escaping SMBH to GW recoil as opposed to three body interactions between an SMBH binary and a third SMBH is also not trivial, as demonstrated in the recent observations presented in \citet{vandokkum2023}.

Numerically, the effect of recoiling SMBHs on the formation of cores in massive elliptical galaxies has been 
primarily studied using collisionless simulations. Early work by \citet{boylan2004}
found that a recoiling SMBH induced a core in the stellar density profile, particularly when the SMBH remained bound to the bulge of the host galaxy and had a recoil velocity comparable to the bulge velocity dispersion. In another early study 
\citet{merritt2004} also found that recoiling SMBHs facilitated the formation of cores
and found that ejected SMBHs settled roughly in one orbital period in spherical potentials, with this `infall time' increasing when more realistic triaxial potentials were considered. 
In \citet{gualandris2008} the motion of a settling SMBH in a stellar potential representative of a galaxy merger remnant was found to occur in three stages: first, the SMBH oscillates with decreasing amplitude due to energy dissipation by dynamical friction.
Next, both the SMBH and stellar core oscillate about their common centre of mass when the radial excursion of the SMBH is less than the core radius of the stellar system.
Finally, the SMBH settles into thermal equilibrium with the surrounding stars.
During these three stages, the stellar density core is enhanced by the motion of the SMBH, producing a mass deficit in some cases up to five times the SMBH mass (see also \citealt{merritt2004,boylan2004,partmann2024}).
However, a more recent study by \citet{nasim2021b} on the contribution to stellar core expansion through GW recoil of an SMBH found intriguingly that the greatest contribution to the stellar mass deficit is due to the first excursion of the SMBH following its ejection, with subsequent oscillations only providing a minor contribution to the growth of the stellar core. 
%This result is in tension with the work of \citet{gualandris2008}.

%The effect of recoiling SMBHs have been studied previously, with some of the first explorations presented in \citet{boylan2004} and \citet{merritt2004}.
%In the work of \citet{boylan2004}, a recoiling SMBH was found to induce a core in the stellar density profile, particularly when the SMBH remained bound to the bulge of the host galaxy and had a recoil velocity comparable to the bulge velocity dispersion.
%A recoiling SMBH was also found to induce a stellar density core in the work of \citet{merritt2004}.
%Additionally, the time for a recoiling SMBH to settle after being ejected in a spherical potential was found to be roughly one orbital period when the recoil velocity was less than the velocity dispersion of the bulge.
%This, which are more representative of a galaxy merger remnant than the spherical case.

%More recently, the contribution of stellar core expansion through GW recoil of an SMBH has been considered by \citet{nasim2021b}.
%Intriguingly, the authors find that greatest contribution to the stellar mass deficit is due to the first excursion of the SMBH following its ejection, with subsequent oscillations only providing a minor contribution to the growth of the stellar core.
%This result is in tension with the work of \citet{gualandris2008}.

%The effect of recoiling SMBHs have been studied previously, with some of the first explorations presented in \citet{boylan2004} and \citet{merritt2004}.
%In the work of \citet{boylan2004}, a recoiling SMBH was found to induce a core in the stellar density profile, particularly when the SMBH remained bound to the bulge of the host galaxy and had a recoil velocity comparable to the bulge velocity dispersion.
%A recoiling SMBH was also found to induce a stellar density core in the work of \citet{merritt2004}.
%Additionally, the time for a recoiling SMBH to settle after being ejected in a spherical potential was found to be roughly one orbital period when the recoil velocity was less than the velocity dispersion of the bulge.
%This `infall time' was found to increase when considering triaxial potentials, which are more representative of a galaxy merger remnant than the spherical case.

%In \citet{gualandris2008}, a stellar potential representative of a galaxy merger remnant with SMBH binary core scouring is used to investigate the motion of an SMBH ejected through GW recoil.
%The motion of a settling SMBH was found to occur in three stages: first, the SMBH oscillates with decreasing amplitude due to energy dissipation by dynamical friction.
%Next, both the SMBH and stellar core oscillate about their common centre of mass when the radial excursion of the SMBH is less than the core radius of the stellar system.
%Finally, the SMBH settles into thermal equilibrium with the surrounding stars.
%During these three stages, the stellar density core is enhanced by the motion of the SMBH (in agreement with \citet{merritt2004} and \citet{boylan2004}), with the mass deficit in some cases reaching up to five times the SMBH mass being produced.

%Work by \citet{blecha2008} considered the effect of recoiling SMBHs in an analytical model of an isolated galaxy comprised of a multicomponent system of a stellar bulge and gaseous disc.
%An interesting result was the mass growth of the ejected SMBH as it settled at the centre of the potential, with the mass growth accumulating to some 10\% of the initial SMBH mass, irrespective of the recoil velocity.
%Recoiling SMBHs in gas-rich systems was further explored in \citet{blecha2011} using hydrodynamic simulations of isolated galaxy mergers.
%In this work, seed SMBHs were allowed to grow through gas accretion prior to merging, with the majority of final SMBH masses at the time of merger sampling the $10^7\,\Msun \lesssim M_\bullet \lesssim 10^8\,\Msun$ mass regime.

%The impact of recoiling SMBHs on the growth and dynamics of SMBHs in hydrodynamical simulations have also been studied using both isolated galaxy models consisting of a stellar bulge and a gaseous disc \citep{blecha2008}, as well as in isolated gas-rich galaxy mergers \citep{blecha2011}. %The study of the isolated galaxy model found that 
%the mass growth of the ejected SMBH as it settled at the centre of the potential, with the mass growth accumulating to some 10\% of the initial SMBH mass, irrespective of the recoil velocity. 
%In the merger simulations, the seed SMBHs were allowed to grow through gas accretion prior to merging, with the majority of the final SMBH masses at the time of merger sampling the $10^7\,\Msun \lesssim M_\bullet \lesssim 10^8\,\Msun$ mass regime. The simulations also showed that recoil events typically reduce the lifetimes of bright active galactic nuclei (AGNs), with most kicked AGNs confined to the central few kpc of the galaxy. Whilst the results of \citet{blecha2011} provide valuable insight into the occurrence of variable AGN variability due to an oscillating SMBH plunging through a gaseous disc, we stress that the results are applicable to a different mass regime of galaxies and SMBHs than what we focus on in this work. In addition, the formation of stellar cores which is primarily driven by complex three-body interactions between the SMBHs and the stellar component remains largely unresolved in simulations that include gravitational softening.

Whilst previous studies have convincingly shown that there is a clear case for a relation between SMBH recoil velocity and the final core size carved out by its motion, little attention has been devoted to understanding the statistical impact of SMBH recoil on the properties of the merger remnant galaxy. 
Invoking large SMBH recoils as a mechanism to grow the most massive cores in elliptical galaxies \citep[e.g.][]{nasim2021b,khonji2024} has contributed an important piece to the overall puzzle of explaining current observations, however recoil velocities approaching the escape velocity of the host galaxy are unlikely to be the norm, given that all observed elliptical galaxies host a central SMBH \citep{ferrarese1994,haring2004,kormendy2013,vandenbosch2016}.
The role of mild SMBH recoil velocities remains largely unexplored.
In addition, how a recoiling SMBH impacts the surrounding stellar orbital distribution has remained largely unanswered. 
Orbital modelling techniques, in particular orbit-superposition Schwarzschild modelling, offer a unique bridging between observations and simulations, often revealing a rich understanding of a galaxy's history, and the effect of SMBH interactions \citep[e.g.][]{rottgers2014,neureiter2021,neureiter2023,santucci2022,santucci2023,santucci2024}.

In this study we use the \gadget{} based regularised tree code \ketju{} \citep{rantala2017,rantala2020,mannerkoski2023} to study the effect of recoiling SMBHs on the structure of massive elliptical galaxies. 
Previously, \ketju{} has been used to study the formation of cores through SMBH binary scouring in gas-free simulations, but without the contribution of SMBH recoils \citep{rantala2018,rantala2019,rantala2024}.
Here, we run a large set of gas-free simulations with systematically varying SMBH kick velocities.  
The aim of this work is twofold.
We first aim to determine how the velocity with which the SMBH is kicked affects the mass distribution of the surrounding stellar environment, and the likelihood of the magnitude of these effects. We then aim to determine if a recoiling SMBH imparts a signature on the stellar kinematics of the remnant galaxy, and if such a signature could potentially be observed with current detectors.

%In this study, we wish to consider SMBH recoil by GW emission in massive elliptical galaxies.
%The aim of this work is twofold.
%We first aim to determine how the velocity with which the SMBH is kicked affects the mass distribution of the surrounding stellar environment, and the likelihood of the magnitude of these effects.
%We then aim to determine if a recoiling SMBH imparts a signature on the stellar kinematics of the remnant galaxy, and if such a signature could be observed with current detectors.

This paper is divided as follows: we present the numerical simulations used in our investigation in \autoref{sec:num_sims}.
We discuss the motion of the recoiling SMBHs in \autoref{sec:motion}, and the resulting stellar density profiles of the remnant galaxy in \autoref{sec:density}.
We investigate the distribution of stellar orbits as a function of the kick velocity in \autoref{sec:orbits} and construct mock stellar kinematic observations to be compared with observational data in \autoref{sec:kinematics}.
Finally, we discuss of our results in \autoref{sec:discussion} and present 
our conclusions in \autoref{sec:conclusions}, respectively.

@@

\subsection{Simulation Code}
To investigate the effect of the recoiling SMBH on the host galaxy, and potential observational signatures of this, we run a number of numerical simulations of an idealised galaxy merger setting using the new public version of the \ketju{}\footnote{\url{https://www.mv.helsinki.fi/home/phjohans/group-website/research/ketju/}} code \citep{rantala2017,rantala2020,mannerkoski2023} coupled with \gadget{} \citep{springel2021}.
\ketju{} integrates the dynamics of SMBHs, and stellar particles in a small region thrice the stellar softening length around them (i.e. the \ketju{} region), to high accuracy using the algorithmically regularised integrator \textsc{mstar} \citep{rantala2020}.
The dynamics of stellar particles beyond this small region of high integration accuracy, and all DM particles, is computed with the \gadget{} fast multiple method (FMM) with second order multipoles. 
Additionally, we use hierarchical time integration, which allows for mutually symmetric interactions and manifest momentum conservation. 
In this study, we do not use comoving integration, as we are investigating an idealised system.
\ketju{} also includes post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014} and the fitting formula of \citet{zlochower2015} for recoil kick velocity, making \ketju{} a particularly well-suited code for investigating the consequence of SMBH recoil self-consistently in a galaxy merger environment.

\subsection{Initial Conditions}
We model the merger of two massive elliptical galaxies using gas-free merger simulations.
Our galaxy initial conditions follow that of IC $\gamma$-1.0-BH-6 presented in \citet{rantala2018} but scaled down by a factor of $\sim 3$ in mass, consistent with the method presented in \citet{rantala2019}.
The galaxy is thus represented as an isotropic multicomponent sphere, and consists of a stellar component of total mass $M_\star\sim 1.38\times10^{11}\,\Msun$ embedded within a DM component of total mass $M_\mathrm{DM}=2.5\times10^{13}\,\Msun$.
At the centre an SMBH with mass $M_{\bullet,0}=2.93\times10^{9}\,\Msun$ is placed with zero velocity.
The stellar and DM components each follow a \citet{hernquist1990} profile with a scale radius of $a_\star=3.9\,\mathrm{kpc}$ and $a_\mathrm{DM}=245\,\mathrm{kpc}$, respectively.
The density profile $\rho_i$ for a given component $i$ is given by:
\begin{equation}\label{eq:dehnen}
    \rho_i(r) = \frac{M_i}{2\pi} \frac{a_i}{r (r+a_i)^{3}}.
\end{equation}
We generate the ICs using the distribution function method following \citet{hilz2012} and \citet{rantala2017}, where for each component (stellar and DM) the distribution function $f_i$ is computed using Eddington's formula \citep{binney2008} for each density profile $\rho_i$:
\begin{equation}
    f_i(\mathcal{E}) = \frac{1}{2\sqrt{2}\pi^2} \int_{\Phi_\mathrm{T}=0}^{\Phi_\mathrm{T}=\mathcal{E}} \dnv{2}{\rho_i}{\Phi_\mathrm{T}} \frac{\dd{\Phi_\mathrm{T}}}{\sqrt{\mathcal{E}-\Phi_\mathrm{T}}}.
\end{equation}
Here $\mathcal{E}$ is the relative energy, and $\Phi_\mathrm{T}$ is the total gravitational potential. 
The distribution functions are then sampled with discrete particles, where the mass of a stellar particle is set to $m_\star=5\times10^4\,\Msun$, and the mass of a DM particle is set to $m_\mathrm{DM}=5\times10^6\,\Msun$.
The radial velocity profiles of both stellar and DM particles are ergodic.

As we are interested in the evolution of the galaxy merger \textit{remnant} following SMBH coalescence, the initial conditions are constructed such that at the time of SMBH coalescence (\autoref{ssec:mergers}) the galaxy remnant agrees with local scaling relations.
Specifically, we ensure that our remnant agrees with the half-light -- stellar mass data presented in \citet{sahu2020}, and lies within the $1\sigma$ prediction interval of the SMBH mass -- stellar velocity dispersion data given in \citet{vandenbosch2016}.
